An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-04-26 EconGrowth.Rnw
Further edits on the preliminary sections.

\noindent where $y \equiv Y/Y_\mathrm{0}$, $k \equiv K/K_\mathrm{0}$, and $l \equiv L/L_\mathrm{0}$; with $Y$ represented by GDP, $K$ expressed in currency units, $L$ expressed in workers/year or work-hours/year, time ($t$) measured in years, e being the base of the natural logarithm, and subscripts ``0'' indicating values at an initial year.\footnote{Indexed quantities are represented by lower-case symbols ($y$, $k$, $l$, $e$, $q$, $x$, and $u$) and dimensional quantities are represented by upper-case symbols ($Y$, $K$, $L$, $E$, $Q$, $X$, and $U$).} $A$ is a technological progress term, with $\lambda$ indicating the pace of technological progress (in units of year$^{-1}$). Neoclassical economic growth theory makes no attempt to understand the mechanisms by which technological progress ($A$) affects economic growth, nor does it attempt to describe how technological progress ($A$) might vary from one economy to the next. Mathematically, $A$ is a fitting parameter that aligns the predictions of Equation \ref{eq:CD} with historical data. Typically, the constraint $\alpha + \beta = 1$ is assumed, indicating constant returns to scale.

@@

Some energy economists augment economic growth models with energy being an additional (or the only) factor of production. However, there is little agreement in the literature on how this should be done. Neither the mathematical form of the energy-augmented production function nor the quantification of energy are consistent in the literature. Furthermore, there has been little empirical work done to assess the predictive ability of energy-augmented production functions for different types of economies. In addition, very few studies of the importance of energy in economic growth include rigorous statistical assessments of energy-related parameters. This paper attempts to contribute to the literature by applying statistically rigorous methods along three coordinates of analysis: form of the energy augmented production function, energy quantification, and type of economy.

@@

\noindent where $e$ is indexed energy and e is the base of the natural logarithm. The LINEX model assumes energy ($e$) is the only factor of production. $\rho_k \equiv \frac{k}{(1/2)(l+e)}$ and represents the rate at which capital stock increases relative to the average of labor and energy. $\rho_l \equiv \frac{l}{e}$ and represents the rate at which labor ($l$) increases relative to energy ($e$). Energy ($e$) as a factor of production is enhanced by $A$ if either

\begin{itemize}
\item{capital stock ($k$) increases faster than the average of energy ($e$) and labor ($l$), i.e. if $\rho_k >$ 1, or}
\item{labor ($l$) increases faster than energy ($e$), i.e. if $\rho_l >$ 1.}
\end{itemize}

\noindent According to the LINEX model, an economy that can increase capital stock ($k$) without a commensurate increase in labor ($l$) and energy ($e$) will experience an increase in output ($y$) in excess of its increase of energy consumption ($e$). Similarly, an economy benefits by increasing labor ($l$) without a commensurate increase in energy ($e$). Thus, $\rho_k$ is an efficiency of using labor and energy to make capital available to the economy, and $\rho_l$ is an efficiency of using energy to make labor available to the economy. The parameter $a_0$ represents the overall importance of these efficiencies, and $c_t$ represents the relative importance of the labor efficiency term compared to the capital stock efficiency term. Both $a_0$ and $c_t$ are constant with respect to time.

In contrast to the Cobb-Douglas model (Equations \ref{eq:CD} and \ref{eq:CDe}) and the CES model (Equations \ref{eq:CES} and \ref{eq:CESe}), the LINEX model (Equation \ref{eq:LINEX}) accounts for efficiencies among capital, labor, and energy directly and, therefore, does not employ a time-dependent technical progress term ($A$). \citet{Kummel:2002tx} use the LINEX production function to describe economic output for the economies of the U.S., Japan, and Germany. \citet{Warr:2010ga} compare the LINEX production function with the Cobb-Douglas production function to assess economic growth for Austria, the United Kingdom, the United States, and Japan.

@@

The literature shows three quantifications of energy ($E$) used in energy-augmented production functions: thermal energy ($Q$), exergy ($X$), and useful work ($U$). Thermal energy ($Q$) is the heat content of an energy carrier. Exergy ($X$) is the maximum useful work that can be extracted from an energy carrier by a process that brings the system into equilibrium with a heat reservoir (typically the surrounding environment). Useful work ($U$) includes the direct energy that goes into an economy, including work from machines, human muscles, and animals. Most energy economists have used thermal energy ($Q$) for the energy factor of production ($E$) in energy-augmented production functions, including \citet{Stern:2012ey}, \citet{Froling:2009vo}, \citet{Serrenho:2010vq}, and \citet{Kummel:1980wx,Kummel:1982vy,Kummel:1985vz,Kummel:1989tf,Kummel:2002tx,Kummel:2010vz}. \citet{Warr:2010ga} introduced the idea of using exergy ($X$) and useful work ($U$) in a production function to account for efficiency gains in energy production and consumption over the last century.

We know of no statistically-rigorous comparisons of the predictive ability of each type of energy quantification in the various types of production functions discussed in Section \ref{sec:Prod_Func_Forms}. We analyze all three quantifications of energy ($Q$, $X$, and $U$) for developed economies and thermal energy ($Q$) and exergy ($X$) for all other economies.

@@

In most energy-economic growth literature, economists assess developed economies from North America or Europe. Examples include \citet{Kummel:1980wx} applying LINEX to describe the West German economy, \citet{Serrenho:2010vq} applying Cobb-Douglas variants to EU-15 countries, \citet{VanderWerf:2008tc} applying CES nesting variants to the 12 OECD countries, \citet{Lecca:2011kr} applying CES nesting variants to the Scottish economy, \citet{Stern:2012ey} applying CES to describe the Swedish economy, and \citet{Warr:2012cg} using LINEX and Cobb-Douglas to describe the Austrian, American, British, and Japanese economies.

In this paper we analyze at least two economies from each of the following categories: developed (technologically advanced with high standards of living), BRICS\footnote{Brazil, Russia, India, China, and South Africa} (rapid economic growth with standards of living below that of developed economies), petrol (petroleum is a substantial source of GDP); and LDCs\footnote{Lesser Developed Countries} (living standards near subsistence levels). (See Table \ref{tab:Economy_Examples}.)

@@

Total employment and average hours worked per year per worker for the United States, the United Kingdom, and Japan for 1980-2011 was obtained from the United States \citet{BLS:2011a}. Total hours worked was calculated by multiplying total employment by average hours worked per worker. ************* Matt ended here. **************** The World Bank's WDI database (WORLD BANK, 2012) provides an employment to population ratio of people fifteen and older, total population, and a percentage of people under fifteen for South Africa, China, Iran, Saudi Arabia, Tanzania, and Zambia for 1991-2011. The population of people fifteen and older was calculated by multiplying the percentage of people under fifteen by total population, and then subtracting this from the total population. Total employment could be calculated by multiplying the employment to population ratio of people fifteen and older by the population of people fifteen and older. The labor input for South Africa, China, Iran, Saudi Arabia, Tanzania, and Zambia is based on total employment, not on total hours worked per year.

@@

We were interested in how the three different types of energy (thermal energy, exergy, and useful work) would affect our econometric growth modeling. The rationale being that exergy, and to a greater extent useful work, would take into account the energy that is being directly used to further economic growth. We examined typical sources of energy for an economy, including: coal; oil and petroleum products; dry natural gas and natural gas plant liquids; nuclear power generation; and renewable energy sources (solar, wind, geothermal, biomass/fuel wood,  hydroelectric generation). In the style of (AYRES AND WARR, 2002, 2005), energy available from human and animal muscle power are also considered. The time series data used for the United States, the United Kingdom, and Japan is from 1980-2011. Time series data used for South Africa, China, Saudi Arabia, Iran, Zambia, and Tanzania is from 1991-2011.

@@

Figure \ref{fig:Historical_Lattice_Graph} indicates the importance of energy quantification for economic growth modeling. Thermal energy ($q$) and exergy ($x$) grow at similar rates, unless there is a significant raw energy carrier substitution (such as China's substitution of ***** for ***** over the last 2 decades). The growth rate of $u$ is significantly different from $q$ and $x$, however. \citet{Warr:2012cg} show that from 1950-1990, useful work ($u$) grew much faster than $q$, $x$, $k$, or $l$ due to efficiency gains in the U.S. economy. The results of \citeauthor{Warr:2012cg} suggest that the time-dependent technological progress term ($A$) in the Cobb-Douglas and CES production functions (Equations \ref{eq:CD}, \ref{eq:CDe}, and \ref{eq:CESe}) may be a proxy for increasing energy efficiency in the U.S. economy.

>>>>

\noindent where $y \equiv Y/Y_\mathrm{0}$, $k \equiv K/K_\mathrm{0}$, and $l \equiv L/L_\mathrm{0}$; with $Y$ represented by GDP, $K$ expressed in currency units, $L$ expressed in workers/year or work-hours/year, time ($t$) measured in years, e being the base of the natural logarithm, and subscripts ``0'' indicating values at an initial year.\footnote{In this paper, indexed quantities are represented by lower-case symbols ($y$, $k$, $l$, $e$, $q$, $x$, and $u$) and dimensional quantities are represented by upper-case symbols ($Y$, $K$, $L$, $E$, $Q$, $X$, and $U$).} $A$ is a technological progress term, with $\lambda$ indicating the pace of technological progress (in units of year$^{-1}$). Neoclassical economic growth theory makes no attempt to understand the mechanisms by which technological progress ($A$) affects economic growth, nor does it attempt to describe how technological progress ($A$) might vary from one economy to the next. Mathematically, $A$ is a fitting parameter that aligns the predictions of Equation \ref{eq:CD} with historical data. Typically, the constraint $\alpha + \beta = 1$ is assumed, indicating constant returns to scale.

@@

Some energy economists augment economic growth models with energy being an additional (or the only) factor of production. However, there is little agreement on how this should be done. Neither the mathematical form of the energy-augmented production function nor the quantification of energy are consistent in the literature. Furthermore, there has been little empirical work done to assess the predictive ability of various energy-augmented production functions for different types of economies. In addition, very few studies of the importance of energy in economic growth include rigorous statistical assessments of energy-related fitting parameters. This paper attempts to contribute to the literature by applying statistically rigorous methods along three coordinates of analysis: form of the energy augmented production function, energy quantification, and type of economy.

@@

\noindent where $e$ is indexed energy and e is the base of the natural logarithm. The LINEX model assumes energy ($e$) is the only factor of production. The parameters $\rho_k$ and $\rho_l$ are defined as

\begin{equation} \label{eq:rho_k}
  \rho_k \equiv \frac{k}{(1/2)(l+e)}
\end{equation}

\noindent and

\begin{equation} \label{eq:rho_l}
  \rho_l \equiv \frac{l}{e}
\end{equation}

\noindent and represent (a) capital stock increase relative to the average of labor and energy and (b) the increase of labor ($l$) relative to energy ($e$), respectively. Energy ($e$) as a the only factor of production is enhanced by $A$ if either

\begin{itemize}
\item{capital stock ($k$) has increased more than the average of energy ($e$) and labor ($l$), i.e. if $\rho_k >$ 1, or}
\item{labor ($l$) has increased more than energy ($e$), i.e. if $\rho_l >$ 1.}
\end{itemize}

\noindent According to the LINEX model, an economy that can increase capital stock ($k$) without a commensurate increase in labor ($l$) and energy ($e$) will experience an increase in output ($y$) in excess of its increase of energy consumption ($e$), because $A$ will be greater than 1. Similarly, an economy benefits by increasing labor ($l$) without a commensurate increase in energy ($e$). Thus, $\rho_k$ is an efficiency of using labor and energy to make capital available to the economy, and $\rho_l$ is an efficiency of using energy to make labor available to the economy. $a_0$ represents the overall importance of these efficiencies, and $c_t$ represents the relative importance of the labor efficiency term compared to the capital stock efficiency term. We treat $a_0$ and $c_t$ as fitting parameters; both are assumed constant with respect to time.

In contrast to both the Cobb-Douglas model (Equations \ref{eq:CD} and \ref{eq:CDe}) and the CES model (Equations \ref{eq:CES} and \ref{eq:CESe}), the LINEX model (Equation \ref{eq:LINEX}) accounts for efficiencies among capital, labor, and energy directly and, therefore, does not employ a time-dependent technical progress term ($A$). \citet{Kummel:2002tx} use the LINEX production function to describe economic output for the economies of the U.S., Japan, and Germany. \citet{Warr:2010ga} compare the LINEX production function with the Cobb-Douglas production function to assess economic growth for Austria, the United Kingdom, the United States, and Japan.

@@

The literature shows three quantifications of energy ($E$) used in energy-augmented production functions: thermal energy ($Q$), exergy ($X$), and useful work ($U$). Thermal energy ($Q$) is the heat content of an energy carrier. Exergy ($X$) is the maximum useful work that can be extracted from an energy carrier by a process that brings the system into equilibrium with a heat reservoir (typically the surrounding environment). Useful work ($U$) includes the direct energy that goes into an economy, including work from machines, human muscles, and animals. Firms pay for thermal energy ($Q$), but useful work ($U$) alters the size and shape of materials, keeps the lights on, and transports goods to consumers. Not all forms of energy are created equal in terms of ability to generate economic activity \citep{Cleveland:2000vl}. Exergy ($X$), and to a greater extent, useful work ($U$) account for the useful energy that goes into generating economic activity, thereby reflecting the ``quality'' of each energy type and the efficiency of machines and devices typically used to convert thermal energy ($Q$) into useful work ($U$).

Most energy economists use thermal energy ($q$) for the energy factor of production ($e$) in energy-augmented production functions, including \citet{Stern:2012ey}, \citet{Froling:2009vo}, \citet{Serrenho:2010vq}, and \citet{Kummel:1980wx,Kummel:1982vy,Kummel:1985vz,Kummel:1989tf,Kummel:2002tx,Kummel:2010vz}. \citet{Warr:2010ga} introduced the idea of using exergy ($x$) and useful work ($u$) in energy-augmentd production functions to account for efficiency gains in energy production and consumption over the last century.

We know of no statistically-rigorous comparisons of the predictive ability of each energy quantification in the various types of production functions discussed in Section \ref{sec:Prod_Func_Forms}. We analyze all three quantifications of energy ($Q$, $X$, and $U$) for developed economies and thermal energy ($Q$) and exergy ($X$) for all other economies.

@@

In the energy-economic growth literature, the focus is almost exclusively on developed economies from North America or Europe. Examples include \citet{Kummel:1980wx} applying LINEX to describe the West German economy, \citet{Serrenho:2010vq} applying Cobb-Douglas variants to EU-15 countries, \citet{VanderWerf:2008tc} applying CES nesting variants to the 12 OECD countries, \citet{Lecca:2011kr} applying CES nesting variants to the Scottish economy, \citet{Stern:2012ey} applying CES to describe the Swedish economy, and \citet{Warr:2012cg} using LINEX and Cobb-Douglas to describe the Austrian, American, British, and Japanese economies.

But, the nearly-complete focus on developed economies misses the role that energy plays in the development process. In this paper we analyze at least two economies from each of the following categories: developed (technologically advanced with high standards of living), BRICS\footnote{Brazil, Russia, India, China, and South Africa} (rapid economic growth with standards of living below that of developed economies), petrol (petroleum is a substantial source of GDP); and LDCs\footnote{Lesser Developed Countries} (living standards near subsistence levels). (See Table \ref{tab:Economy_Examples}.)

@@

Total employment and average hours worked per worker per year for the developed nations (United States, United Kingdom, and Japan) for 1980-2011 were obtained from the United States \citet{BLS:2011a}. Total hours worked per year was calculated by multiplying total employment by average hours worked per worker per year.

The World Bank's World Development Indicators database \citep{WorldBankWDI:2013a} provides the employment to population ratio for people fifteen and older, total population, and a percentage of people under fifteen for other nations (South Africa, China, Iran, Saudi Arabia, Tanzania, and Zambia) for 1991-2011. The population of people under fifteen was subtracted from total population to calculate the population of people fifteen and older. Total employment was calculated by multiplying the employment to population ratio of people fifteen and older by the population of people fifteen and older.

@@

************* Matt ended here. ****************

To account for all energy input to an economy, we summed consumption data for many energy sources for each economy, including: coal; oil and petroleum products; dry natural gas and natural gas plant liquids; nuclear power generation; and renewable energy sources (solar, wind, geothermal, biomass/fuel wood,  hydroelectric generation). In the style of (AYRES AND WARR, 2002, 2005), energy available from human and animal muscle power are also considered. The time series data used for the United States, the United Kingdom, and Japan is from 1980-2011. Time series data used for South Africa, China, Saudi Arabia, Iran, Zambia, and Tanzania is from 1991-2011.

@@

Figure \ref{fig:Historical_Lattice_Graph} indicates the importance of energy quantification for economic growth modeling. Thermal energy ($q$) and exergy ($x$) grow at similar rates, unless there is a significant raw energy carrier substitution in the economy (such as China's rapid increase of coal consumption relative to petroleum since 2000). The growth rate of $u$ is significantly different from $q$ and $x$ for most countries, however. \citet{Warr:2012cg} show that from 1950-1990, useful work ($u$) grew much faster than $q$, $x$, $k$, or $l$ due to efficiency gains in the U.S. economy. The results of \citeauthor{Warr:2012cg} suggest that the time-dependent technological progress term ($A$) in the Cobb-Douglas and CES production functions (Equations \ref{eq:CD}, \ref{eq:CDe}, and \ref{eq:CESe}) may be a proxy for increasing energy efficiency in the U.S. economy.
